Hilbert space approximate Bayesian Gaussian processes using Stan

### 2020-04-09 Paper/manuscript_v2/manuscript_07_01.tex
spell check

\begin{abstract}
Gaussian processes are powerful non-parametric probabilistic models for stochastic functions, however they entail a complexity that is computationally intractable when the number of observations is large, especially when estimated with fully Bayesian methods such as Markov-Chain Monte-Carlo. In this paper, we focus on a novel approach for low-rank approximate Bayesian Gaussian processes, based on a basis functions approximation via Laplace eigenfunctions for stationary covariance functions. The main contribution of this paper is a detailed analysis of the performance and practical implementation of the method in relation to key factors such as the number of basis functions, desired prediction space and smoothness of the function to be learned. Intuitive visualizations and useful recommendations for the values of these factors which help users to improve performance and computation are provided. A diagnosis procedure for whether the chosen values for the number of basis functions and the desired prediction space are adequate to fit to the actual data is also proposed. The proposed method is simple and exhibits an attractive computational complexity due to its linear structure and it is easy to implement in probabilistic programming frameworks. Several illustrative examples of the performance and applicability of the method in the probabilistic programming language Stan are presented together with the underlying Stan model code.
\end{abstract}

@@

The Laplace eigenfunctions can be computed analytically and they are independent of the particular choice of the covariance function including the hyperparameters. While the pre-computation cost of the basis functions is $O(m^2n)$, the computational cost of learning the covariance function parameters is $O(mn+m)$ in every step of the optimizer or sampler. This is a big advantage in terms of speed for iterative algorithms such as Markov chain Monte Carlo (MCMC). Another advantage is the reduced memory requirements of automatic differentation methods used in modern probabilistic programming frameworks, such as Stan \citep{carpenter2017stan}, WinBUGS \citep{lunn2000winbugs} and others. This is because the memory requirements of automatic differentation rather scale with the computational complexity instead of with the usual memory requirements for the posterior density computation \todo{I don't understand the sentence about autodiff and computational complexity} \resp{Res: This sentence scapes me. I think this was written by Aki}. The basis function approach also provides an easy way to apply a non-centered parameterization of GPs, which reduces the posterior dependency between parameters representing the estimated function and the hyperparameters of the covariance function, which further improves MCMC efficiency.

@@

The specific numbe of basis functions  $m$ in the approximation needed to satisfy equation (\ref{eq_diff_covs}) depends on the degree of non-linearity of the function to be estimated, that is on its lengthscale $\ell$, which constitutes a hyperparameter of the GP. The approximation also depends on the boundary $L$ (see equations (\ref{eq_eigenvalue}), (\ref{eq_eigenfunction}), (\ref{eq_eigenvalue_multi}) and (\ref{eq_eigenfunction_multi})), which will affect its accuracy especially near the boundaries. As we will see later on, $L$ will also influence the number of basis functions required in the approximation. In the present paper, we will set $L$ an extension of the desired covariate input domain $\Psi = \max_{i}|x_i|$. Without loss of generality, we can assume $\Psi$ to be symmetric around zero, that is $\Psi=[-S,S] \subset \Omega$. We now define $L$ as
%
\begin{equation}\label{eq_boundary}
L=c \cdot S,
\end{equation}

@@

As stated above, Figures \ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_lscale_vs_J_vs_c_Matern} provide the minimum lengthscale that can be closely approximated given $m$ and $c$. This information serves as a powerful diagnostic tool in determining if the obtained accuracy is acceptable. As the lenghscale $\ell$ controls the "wigglyness" of
the function, it strongly influences the difficulty of obtaining accurate
inference about the function from the data. Basically, if the lengthscale estimate is accurate, 
we can expect the HSGP approximation to be accurate as well. Thus, having obtained an estimate $\hat{\ell}$ of $\ell$ from the HSGP model based on prespecified $m$ and $c$, we can check whether or not $\hat{\ell}$ exceeds the minimum lengthscale provided in Figure \ref{fig6_lscale_vs_J_vs_c} or \ref{fig6_lscale_vs_J_vs_c_Matern} (depending on which kernel is used). If $\hat{\ell}$ exceeds this recommended minimum lengthscale, the approximation should be close enough. If, however, $\hat{\ell}$ does not exceed recommended minimum lengthscale, the approximation may be inaccurate and $m$ should be increased or $c$ decreased. We may also use this diagnostic in a iterative procedure.
Starting from some initial guess of $\ell$, we can choose initial values for $m$ and $c$ and fit an HSGP model, then check the approximation accuracy, and, if not accurate enough because the estimated $\hat{\ell}$ is below the minimum lengthscale, repeat the process while increasing $m$ or decreasing $c$.
As mentioned earlier, $c$ cannot be decreased as much as desired because it is restricted by the lengthscale, so increasing $m$ may usually the preferred approach.

@@

Figures \ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_lscale_vs_J_vs_c_Matern} were build for a GP with a unidimensional covariance function, which result in a surface depending on three variables, $m$, $c$ and $\ell$. An equivalent figure for a GP model with a two-dimensional covariance function would result in a surface depending on four variables, $m$, $c$, $\ell_1$ and $\ell_2$, which is more difficult to be graphically represented. More precisely, in the multi-dimensional case, whether the approximation is close enough might depend only on the ratio between wiggliness in every dimensions. For instance, in the two-dimensional case, it would depend on the ratio between $\ell_1$ and $\ell_2$ and could be graphically represented. Future research will focus on building useful graphs or analytical models that provide these relations in multi-dimensional cases. However, as an approximation, we can use the unidimensional GP conclusions in Figures \ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_lscale_vs_J_vs_c_Matern} to check the accuracy by analyze individually the different dimensions of a multidimensional GP model.

@@

Comparing the accuracy of the lengthscale in Figure 
\ref{fig8_Tlscale_vs_Elscale} to the RMSE in Figure \ref{fig9_MSE_varing_lscale},
we see that they agree closely with each other for medium lengthscales.
That is, a good estimation of the lengthscale implies a
small RMSE. This is no longer true for very small or large lengthscales.
In small lengthscales, even very small inaccuracies 
may have a strong influence on the posteriors predictions and thus on the RMSE.
In large lengthscales, larger inaccuracies change the posterior predictions
only little and may thus not yield large RMSEs.
The dashed black line in Figure \ref{fig8_Tlscale_vs_Elscale} represents the 
minimum lengthscale that can be closely approximated under the given condition, according to the results presented in Figure \ref{fig6_lscale_vs_J_vs_c}. 
We observe that whenever the estimated lengthscale exceeds the minimially
estimable lengthscale, the RMSE of the posterior predictions is small
(see Figure \ref{fig9_MSE_varing_lscale}).
Conversely, when the estimated lengthscale is smaller than the minimally
estimable one, the RMSE becomes very large.

@@

In the online supplemental material, some other case studies are presented. From those examples, it can be seen how computation time of the HSGP model increases rapidly with the number of input dimensions ($D$) since the number of basis functions in the approximation increases exponentially with $D$ (see equation (\ref{eq_m_multi})). Even though, in a bivariate input space, the computation time increases significantly with $D$, the HSGP model works significantly faster than the regular GP for most of the non-linear $2D$ functions (even highly wiggly functions; see Figures B.3-right and C.3 in the online material). However, HSGPs tend to be slower than exact GPs for $D>3$ with a relatively low number of basis functions ($m \gtrsim 5$), as well as even for $D=3$ with a moderate high number of basis functions ($m \gtrsim 20$; see Figure C.3 in the online material). In all of the investigated cases, choosing the optimal boundary factor in the HSGP approximation reduces the number of required basis functions noticably (see Figures A.3, B.3-left and C.2 in the online material) and therefore also reduces computational time drastically in particular in multivariate input spaces.

@@

The developed approximate GPs can be easily applied as modular components in probabilistic programming frameworks such as Stan in both Gaussian and non-Gaussian observational models. Using several simulated and real datasets, we have demonstrated the practical applicability and improved sampling effiency, as compared to regular GPs, of the developed method.
The main drawback of the approach is that its computational complexity scales exponentially with the number of input dimensions. Hence, choosing optimal values for the number of basis functions and the boundary factor, using the
 recommendations and diagnostics provided in Figures \ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_lscale_vs_J_vs_c_Matern}, is essential to avoid a excessive computational time especially in multivariate input spaces. However, in practice, input dimensionalities larger than 3 start to be quite computationally demanding even for moderately wiggly functions and few basis functions per input dimension. In these high dimensional cases, the proposed approxiamte GP methods may still be used for low-dimensional components in an additive modeling scheme but without modeling very high dimensional interactions.

@@

\noindent the covariance $k(\tau)$ in equation (\ref{ch5_eq_cov_periodic_taylor_approx}) can be re-writting as
%
\begin{equation} \label{ch5_eq_cov_periodic_taylor_approx_2}
k(x,x')= \sigma^2 \Big( \sum_{j=0}^{J} \tilde{q}_j^2 \cos(j\omega_0 x)  \cos(j\omega_0 x') + \sum_{j=1}^{J} \tilde{q}_j^2 \sin(j\omega_0 x) \sin(j\omega_0 x') \Big).
\end{equation}

>>>>

\begin{abstract}
Gaussian processes are powerful non-parametric probabilistic models for stochastic functions, however they entail a complexity that is computationally intractable when the number of observations is large, especially when estimated with fully Bayesian methods such as Markov chain Monte Carlo. In this paper, we focus on a novel approach for low-rank approximate Bayesian Gaussian processes, based on a basis functions approximation via Laplace eigenfunctions for stationary covariance functions. The main contribution of this paper is a detailed analysis of the performance and practical implementation of the method in relation to key factors such as the number of basis functions, desired prediction space and smoothness of the function to be learned. Intuitive visualizations and useful recommendations for the values of these factors which help users to improve performance and computation are provided. A diagnosis procedure for whether the chosen values for the number of basis functions and the desired prediction space are adequate to fit to the actual data is also proposed. The proposed method is simple and exhibits an attractive computational complexity due to its linear structure and it is easy to implement in probabilistic programming frameworks. Several illustrative examples of the performance and applicability of the method in the probabilistic programming language Stan are presented together with the underlying Stan model code.
\end{abstract}

@@

The Laplace eigenfunctions can be computed analytically and they are independent of the particular choice of the covariance function including the hyperparameters. While the pre-computation cost of the basis functions is $O(m^2n)$, the computational cost of learning the covariance function parameters is $O(mn+m)$ in every step of the optimizer or sampler. This is a big advantage in terms of speed for iterative algorithms such as Markov chain Monte Carlo (MCMC). Another advantage is the reduced memory requirements of automatic differentiation methods used in modern probabilistic programming frameworks, such as Stan \citep{carpenter2017stan}, WinBUGS \citep{lunn2000winbugs} and others. This is because the memory requirements of automatic differentiation rather scale with the computational complexity instead of with the usual memory requirements for the posterior density computation \todo{I don't understand the sentence about autodiff and computational complexity} \resp{Res: This sentence escapes me. I think this was written by Aki}. The basis function approach also provides an easy way to apply a non-centered parameterization of GPs, which reduces the posterior dependency between parameters representing the estimated function and the hyperparameters of the covariance function, which further improves MCMC efficiency.

@@

The specific number of basis functions  $m$ in the approximation needed to satisfy equation (\ref{eq_diff_covs}) depends on the degree of non-linearity of the function to be estimated, that is on its lengthscale $\ell$, which constitutes a hyperparameter of the GP. The approximation also depends on the boundary $L$ (see equations (\ref{eq_eigenvalue}), (\ref{eq_eigenfunction}), (\ref{eq_eigenvalue_multi}) and (\ref{eq_eigenfunction_multi})), which will affect its accuracy especially near the boundaries. As we will see later on, $L$ will also influence the number of basis functions required in the approximation. In the present paper, we will set $L$ an extension of the desired covariate input domain $\Psi = \max_{i}|x_i|$. Without loss of generality, we can assume $\Psi$ to be symmetric around zero, that is $\Psi=[-S,S] \subset \Omega$. We now define $L$ as
%
\begin{equation}\label{eq_boundary}
L=c \cdot S,
\end{equation}

@@

As stated above, Figures \ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_lscale_vs_J_vs_c_Matern} provide the minimum lengthscale that can be closely approximated given $m$ and $c$. This information serves as a powerful diagnostic tool in determining if the obtained accuracy is acceptable. As the lengthscale $\ell$ controls the "wigglyness" of
the function, it strongly influences the difficulty of obtaining accurate
inference about the function from the data. Basically, if the lengthscale estimate is accurate, 
we can expect the HSGP approximation to be accurate as well. Thus, having obtained an estimate $\hat{\ell}$ of $\ell$ from the HSGP model based on prespecified $m$ and $c$, we can check whether or not $\hat{\ell}$ exceeds the minimum lengthscale provided in Figure \ref{fig6_lscale_vs_J_vs_c} or \ref{fig6_lscale_vs_J_vs_c_Matern} (depending on which kernel is used). If $\hat{\ell}$ exceeds this recommended minimum lengthscale, the approximation should be close enough. If, however, $\hat{\ell}$ does not exceed recommended minimum lengthscale, the approximation may be inaccurate and $m$ should be increased or $c$ decreased. We may also use this diagnostic in a iterative procedure.
Starting from some initial guess of $\ell$, we can choose initial values for $m$ and $c$ and fit an HSGP model, then check the approximation accuracy, and, if not accurate enough because the estimated $\hat{\ell}$ is below the minimum lengthscale, repeat the process while increasing $m$ or decreasing $c$.
As mentioned earlier, $c$ cannot be decreased as much as desired because it is restricted by the lengthscale, so increasing $m$ may usually the preferred approach.

@@

Figures \ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_lscale_vs_J_vs_c_Matern} were build for a GP with a unidimensional covariance function, which result in a surface depending on three variables, $m$, $c$ and $\ell$. An equivalent figure for a GP model with a two-dimensional covariance function would result in a surface depending on four variables, $m$, $c$, $\ell_1$ and $\ell_2$, which is more difficult to be graphically represented. More precisely, in the multi-dimensional case, whether the approximation is close enough might depend only on the ratio between wigglyness in every dimensions. For instance, in the two-dimensional case, it would depend on the ratio between $\ell_1$ and $\ell_2$ and could be graphically represented. Future research will focus on building useful graphs or analytical models that provide these relations in multi-dimensional cases. However, as an approximation, we can use the unidimensional GP conclusions in Figures \ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_lscale_vs_J_vs_c_Matern} to check the accuracy by analyze individually the different dimensions of a multidimensional GP model.

@@

Comparing the accuracy of the lengthscale in Figure 
\ref{fig8_Tlscale_vs_Elscale} to the RMSE in Figure \ref{fig9_MSE_varing_lscale},
we see that they agree closely with each other for medium lengthscales.
That is, a good estimation of the lengthscale implies a
small RMSE. This is no longer true for very small or large lengthscales.
In small lengthscales, even very small inaccuracies 
may have a strong influence on the posteriors predictions and thus on the RMSE.
In large lengthscales, larger inaccuracies change the posterior predictions
only little and may thus not yield large RMSEs.
The dashed black line in Figure \ref{fig8_Tlscale_vs_Elscale} represents the 
minimum lengthscale that can be closely approximated under the given condition, according to the results presented in Figure \ref{fig6_lscale_vs_J_vs_c}. 
We observe that whenever the estimated lengthscale exceeds the minimally
estimable lengthscale, the RMSE of the posterior predictions is small
(see Figure \ref{fig9_MSE_varing_lscale}).
Conversely, when the estimated lengthscale is smaller than the minimally
estimable one, the RMSE becomes very large.

@@

In the online supplemental material, some other case studies are presented. From those examples, it can be seen how computation time of the HSGP model increases rapidly with the number of input dimensions ($D$) since the number of basis functions in the approximation increases exponentially with $D$ (see equation (\ref{eq_m_multi})). Even though, in a bivariate input space, the computation time increases significantly with $D$, the HSGP model works significantly faster than the regular GP for most of the non-linear $2D$ functions (even highly wiggly functions; see Figures B.3-right and C.3 in the online material). However, HSGPs tend to be slower than exact GPs for $D>3$ with a relatively low number of basis functions ($m \gtrsim 5$), as well as even for $D=3$ with a moderate high number of basis functions ($m \gtrsim 20$; see Figure C.3 in the online material). In all of the investigated cases, choosing the optimal boundary factor in the HSGP approximation reduces the number of required basis functions noticeably (see Figures A.3, B.3-left and C.2 in the online material) and therefore also reduces computational time drastically in particular in multivariate input spaces.

@@

The developed approximate GPs can be easily applied as modular components in probabilistic programming frameworks such as Stan in both Gaussian and non-Gaussian observational models. Using several simulated and real datasets, we have demonstrated the practical applicability and improved sampling efficiency, as compared to regular GPs, of the developed method.
The main drawback of the approach is that its computational complexity scales exponentially with the number of input dimensions. Hence, choosing optimal values for the number of basis functions and the boundary factor, using the
 recommendations and diagnostics provided in Figures \ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_lscale_vs_J_vs_c_Matern}, is essential to avoid a excessive computational time especially in multivariate input spaces. However, in practice, input dimensionalities larger than 3 start to be quite computationally demanding even for moderately wiggly functions and few basis functions per input dimension. In these high dimensional cases, the proposed approximate GP methods may still be used for low-dimensional components in an additive modeling scheme but without modeling very high dimensional interactions.

@@

\noindent the covariance $k(\tau)$ in equation (\ref{ch5_eq_cov_periodic_taylor_approx}) can be re-writing as
%
\begin{equation} \label{ch5_eq_cov_periodic_taylor_approx_2}
k(x,x')= \sigma^2 \Big( \sum_{j=0}^{J} \tilde{q}_j^2 \cos(j\omega_0 x)  \cos(j\omega_0 x') + \sum_{j=1}^{J} \tilde{q}_j^2 \sin(j\omega_0 x) \sin(j\omega_0 x') \Big).
\end{equation}
